---
title: "Data Manipulation"
format: pdf
editor: visual
---

```{r}
## you can set your working directory in this chunk 

#setwd("/Users/namigabbasov/Desktop/R-Data-Carpentry")
```

```{r}
## libraries 

#install.packages("tidyverse")
#install.packages("magrittr")

library(tidyverse)
library(magrittr)
```

```{r}
## import data
surveys<-read_csv("https://raw.githubusercontent.com/UnitForDataScience/RWorkshop/main/portal_data_joined.csv")
```


# Data manipulation using dplyr

We’re going to learn some of the most common dplyr functions:

select(): subset columns
filter(): subset rows on conditions
mutate(): create new columns by using information from other columns
group_by() and summarize(): create summary statistics on grouped data
arrange(): sort results
count(): count discrete values

## Selecting columns 
```{r}
select(surveys, plot_id, species_id, weight)
```


## Filtering rows
```{r}
select(surveys, -record_id, -species_id)
filter(surveys, year == 1995)
```


## Two or More Fucntions at the Same Time: Pipes
```{r}
## Intermediate steps 

surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)

## nested functions
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)


## pipes
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

## Alternative New Pipe Operator 
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```


```{r}
## Assign it to a new object(surveys_sml) 

surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

surveys_sml
```
## Mutate: Make new variables 
```{r}
surveys %>%
  mutate(weight_kg = weight / 1000)
```


```{r}
## create two columns at the same time

surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
```

```{r}
## you can add other functions to Pipe as well 
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

```{r}
## We can also filter missing observations before making a new column
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```


## Operations on Groups: The group_by() and summarize() functions
```{r}
## We group the data by "sex" column
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

```{r}
## You can also group by multiple columns 
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  tail()
```

```{r}
## Filter out NAs before grouping 
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```
```{r}
## Add print function to see the first 15 observations 
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)
```


```{r}
## summarize multiple columns 
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight))
```

## arrange(): sort results
```{r}
surveys|>
  arrange(hindfoot_length)
```
```{r}
## Let's make it in descending order
surveys|>
  arrange(desc(hindfoot_length))
```

```{r}
## we can also arrange after other data manipulation operations 
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```
## Counting
```{r}
## let's count how many categories we have in "sex" variable 
surveys %>%
    count(sex)
```
```{r}
## grouping by "sex" and summarizing will give the same result. 
surveys %>%
    group_by(sex) %>%
    summarize(count = n())
```

```{r}
## The sort = TRUE argument sorts the resulting counts in descending order.
surveys %>%
    count(sex, sort = TRUE)
```

```{r}
## count combination of factors 
surveys %>%
  count(sex, species)
```
```{r}
## arrange in descending order for species(alphabetically)
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```
```{r}
## arrange in descending order for n 
surveys %>%
  count(sex, species, sort = TRUE) 
```



















# Reshaping with pivot_wider and pivot_longer

1. Each variable has its own column
2. Each observation has its own row
3. Each value must have its own cell
4. Each type of observational unit forms a table
```{r}
## First, let's make a new dataframe from our "Surveys" dataset
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(plot_id, genus) %>%
  summarize(mean_weight = mean(weight))
```
```{r}
## check variables 
str(surveys_gw)
```


## Pivot_wider

```{r}
surveys_wide <- surveys_gw %>%
  pivot_wider(names_from = genus, values_from = mean_weight)

str(surveys_wide)
```

```{r}
## fill in missing observations 
surveys_gw %>%
  pivot_wider(names_from = genus, values_from = mean_weight, values_fill = 0) %>%
  head()
```


## Pivot_longer

```{r}
surveys_long <- surveys_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = -plot_id)

str(surveys_long)
```
```{r}
## look at first observations 
surveys_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = -plot_id) %>%
  head()
```
# Exporting data
```{r}
## Let's make a new data 


surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex



## Extract the most common species_id
species_counts <- surveys_complete %>%
    count(species_id) %>%
    filter(n >= 50)

## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)
```


```{r}
# Export the data final data 
write_csv(surveys_complete, file = "surveys_complete.csv")
```

```{r}
## export it as excel 

#install.packages("writexl")

library(writexl)
write_xlsx(x= surveys_complete, path="surveys_complete.xlsx", col_names=TRUE)
```




